---
title: "DA3_1_koncz_tamas"
author: "Tamas Koncz"
date: '2017 november 12 '
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(ggplot2)

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

#### 1. Filter the data to the city of your choice and other characteristics (stars, accomodation type) . Describe the distribution of the price and distance variables. Comment on graphs. (1-2 sentences)

Loading data & filtering:

```{r, error = FALSE}
hotels <- fread('hotels_all_nov21.csv')

dt <- hotels[accommodation_type %in% c('Hotel', 'Hostel'), ]
dt <- dt[stars >= 2, ]
dt <- dt[city == 'Barcelona', ]
```

Let's look at the distribution of the 'price' and 'distance' variables (codes are hidden with 'echo = FALSE' for better readability of the final document):

```{r, error = FALSE, warning = FALSE, echo = FALSE, message = FALSE, fig.width = 10}

hist_price <- ggplot(dt, aes(price)) + geom_histogram(binwidth = 10) +
  labs(x = "Hotel price (EUR)") +
  geom_vline(xintercept = mean(dt$price), color = 'red') +
  geom_text(aes(x = mean(dt$price), y = 55), label ='mean', color = 'red', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = median(dt$price), color = 'blue') +
  geom_text(aes(x = median(dt$price), y = 55), label ='median', color = 'blue', angle = 90, vjust = 1.0) 
  

hist_distance <- ggplot(dt, aes(distance)) + geom_histogram(binwidth = .25) +
  labs(x = "Distance to city center (km)") +
  geom_vline(xintercept = mean(dt$distance), color = 'red') +
  geom_text(aes(x = mean(dt$distance), y = 65), label ='mean', color = 'red', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = median(dt$distance), color = 'blue') +
  geom_text(aes(x = median(dt$distance), y = 65), label ='median', color = 'blue', angle = 90, vjust = 1.0)


multiplot(hist_price, hist_distance, cols = 2)
```

As we can see, both are skewed to the right, with long left tails.

```{r, error = FALSE}
p <- ggplot(dt, aes(distance, price)) + geom_point() + 
  geom_smooth() + 
  scale_y_continuous(expand = c(0, 0), limits = c(0,500))
p
p + facet_wrap(~stars)
```


