---
title: "DA3_2"
author: "Tamas Koncz"
date: '2017 november 18 '
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.width = 10)
knitr::opts_chunk$set(fig.height = 3)
knitr::opts_chunk$set(fig.align='center')


library(ggplot2)
library(data.table)
library(wbstats)
library(tidyverse)
library(lspline)
library(pander)


# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

#### Download cross-country data on life expectancy and GDP per capita. “GDP per capita, PPP (constant)” and “Life expectancy at birth (total)”

#### 1. Delete unnecessary columns and save a csv file with three columns only: country name, life expectancy and GDP per capita. Keep countries with non-missing values for life expectancy and GDP per capita. Document what you do. 

First, we'll use the WB API for getting our data set to R.


```{r data_collection}

x <- wb(country = "countries_only", indicator = c("NY.GDP.PCAP.PP.KD", "SP.DYN.LE00.IN", "SP.POP.TOTL") , startdate = 2015, enddate = 2015)
countries <- data.table(x)

```

Once we have the data loaded, we format the data into a workable format.
We are making four transformations:

  1. Separate columns created for each of our variables
  
  2. Columns are renamed to short names for easier coding
  
  3. Rows with NA values are removed
  
  4. Population is redefined as population in millions for easier handling
  
As a last step, we save our data table to a csv file as requested in the exercise.

```{r data_prep_2}
dt <- countries[, c('value', 'country', 'indicator')]
dt <- spread(dt, key = 'indicator', value = 'value', fill = NA)  

#saving string names for later use (charts, ...)
s_country = colnames(dt)[1]
s_gdp = "GDP / capita, $" ##colnames(dt)[2]
s_life_exp = "Life exp. at birth, yrs" ##colnames(dt)[3]
s_pop = "Population, million" ##colnames(dt)[4]
s_lnpop = "log(Population, million)"
s_lngdp = "log(GDP per capita, $)"

setnames(dt, old = colnames(dt), new = c('country', 'gdp', 'life_exp', 'pop'))

dt <- dt[is.na(gdp) == FALSE, ]
dt <- dt[is.na(pop) == FALSE, ]
dt <- dt[is.na(life_exp) == FALSE, ]

dt[, pop := pop / 1000000]
dt[, lngdp := log(gdp)]

write.csv(x = dt, file = 'WD_Data_Filtered.csv')
```

Before we create our models, let's explore our data set visually.
First, using histograms to visualize single-variable distributions:


```{r data_viz_1, echo = FALSE}
p1 <- ggplot(data = dt, aes(gdp)) + geom_histogram(binwidth = 2500, fill = 'orange') + 
  scale_x_continuous(labels = scales::comma) + 
  labs(x = s_gdp) +
  geom_text(aes(x = median(dt$gdp), y = 35), label ='median', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = median(dt$gdp), color = 'blue') +
  geom_text(aes(x = mean(dt$gdp), y = 35), label ='mean', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = mean(dt$gdp), color = 'blue')


p2 <- ggplot(data = dt, aes(lngdp)) + geom_histogram(binwidth = .25, fill = 'orange') +
  labs(x = s_lngdp) +
  geom_text(aes(x = median(dt$lngdp), y = 12.5), label ='median', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = median(dt$lngdp), color = 'blue') +
  geom_text(aes(x = mean(dt$lngdp), y = 12.5), label ='mean', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = mean(dt$lngdp), color = 'blue')

multiplot(p1, p2, cols = 2)

p3 <- ggplot(data = dt, aes(life_exp)) + geom_histogram(binwidth = 5, fill = 'orange') + 
  labs(x = s_life_exp) +
  geom_text(aes(x = median(dt$life_exp), y = 45), label ='median', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = median(dt$life_exp), color = 'blue') +
  geom_text(aes(x = mean(dt$life_exp), y = 45), label ='mean', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = mean(dt$life_exp), color = 'blue')

p4 <- ggplot(data = dt, aes(pop)) + geom_histogram(binwidth = 25, fill = 'orange') + 
  scale_x_continuous(labels = scales::comma) + 
  labs(x = s_pop) +
  geom_text(aes(x = median(dt$pop), y = 90), label ='median', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = median(dt$pop), color = 'blue') +
  geom_text(aes(x = mean(dt$pop), y = 90), label ='mean', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = mean(dt$pop), color = 'blue')

p5 <- ggplot(data = dt, aes(log(pop))) + geom_histogram(fill = 'orange') + 
  scale_x_continuous(labels = scales::comma) + 
  labs(x = paste("log(", s_pop, ")", sep = "")) +
  geom_text(aes(x = median(log(dt$pop)), y = 15), label ='median', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = median(log(dt$pop)), color = 'blue') +
  geom_text(aes(x = mean(log(dt$pop)), y = 15), label ='mean', color = 'blue', angle = 90, vjust = 1.0) +
  geom_vline(xintercept = mean(log(dt$pop)), color = 'blue')

multiplot(p3, p4, p5, cols = 3)
```
  
  
Then, we'll take a look at multi-varite distributions with scatterplots (inc. default loess lines).
(Bubble sizes show the size of the population)  


```{r data_viz_2, echo = FALSE, fig.height = 6}
gp <- geom_point(size = log(dt$pop) / 1.5, color = 'orange')
gs <- geom_smooth(method = 'loess')

p6 <- ggplot(data = dt, aes(lngdp, life_exp)) + gp + 
  gs + 
  labs(x = s_lngdp, y = s_life_exp)

p8 <- ggplot(data = dt, aes(lngdp, pop)) + geom_point(size = log(dt$pop) / 1.5, color = 'orange') + 
  geom_smooth(method = 'loess') + 
  labs(x = s_lngdp, y = s_pop)

p7 <- ggplot(data = dt, aes(life_exp, pop)) +  gp + 
  gs +
  labs(x = s_life_exp, y = s_pop)

p9 <- ggplot(data = dt, aes(lngdp, log(pop))) + gp + 
  gs + 
  labs(x = s_lngdp, y = s_lnpop)

multiplot(p6, p7, p8, p9, cols = 2)
```



#### 2. Estimate a lowess regression of life expectancy on ln gdp per capita. Estimate a linear regression of life expectancy on GDP per capita that best captures the nonlinearity you found (life expectancy on a piecewise linear spline or a polynomial in the explanatory variable). Argue for your choice. Report the coefficient estimates as well as their confidence interval, interpret and visualize the results. 


Below code was used to create our different models:  


```{r models, echo = FALSE}
##lowess
lowess_model <- loess(life_exp ~ lngdp, data = dt)
dt$lowess_pred <- predict(lowess_model)
##summary(lowess_model)

r2_lowess <- var(dt$lowess_pred) / var(dt$life_exp)

##linear regression
lm_ln_model <- lm(life_exp ~ lngdp, data = dt)
dt$lm_ln_pred <- predict(lm_ln_model)

r2_lm_ln <- summary(lm_ln_model)$r.squared
##summary(lm_ln_model)
formula_lm_ln = paste(s_life_exp,"~", round(coefficients(lm_ln_model)[1], 2), 
                      "+", round(coefficients(lm_ln_model)[2], 2), "*", s_lngdp, sep = " ")

##significant on all usual confidence levels!

lm_model <- lm(life_exp ~ gdp, data = dt)
dt$lm_pred <- predict(lm_model)

r2_lm <- summary(lm_model)$r.squared
##summary(lm_model)
formula_lm = paste(s_life_exp,"~", round(coefficients(lm_model)[1], 2), 
                      "+", format(coefficients(lm_model)[2], digits = 2), "*", s_gdp, sep = " ")
##significant on all usual confidence levels!

##spline
knot = 10.5
spline_model <- lm(life_exp ~ lspline(lngdp, knot), data = dt)
dt$spline_pred <- predict(spline_model)

r2_spline <- summary(spline_model)$r.squared
##summary(spline_model)
formula_spline = paste(s_life_exp,"~", round(coefficients(spline_model)[1], 2), 
                      "+\n    ", round(coefficients(spline_model)[2], 2), "*", s_lngdp, "(<10.5)", 
                      "+\n    ", round(coefficients(spline_model)[3], 2), "*", s_lngdp, "(>10.5)", sep = " ")


## polynomial
dt$lngdp_sq <- dt$lngdp ^ 2
dt$lngdp_cub <- dt$lngdp ^ 3

lm_sq_model <- lm(life_exp ~ lngdp + lngdp_sq, data = dt)
dt$lm_sq_pred <- predict(lm_sq_model)

r2_lm_sq <- summary(lm_sq_model)$r.squared
##summary(lm_sq_model)
formula_lm_sq = paste(s_life_exp,"~", round(coefficients(lm_sq_model)[1], 2), 
                      "+\n    ", round(coefficients(lm_sq_model)[2], 2), "*", s_lngdp, 
                      "+\n    ", round(coefficients(lm_sq_model)[3], 2), "*", s_lngdp, sep = " ")

lm_cub_model <- lm(life_exp ~ lngdp + lngdp_sq + lngdp_cub, data = dt)
dt$lm_cub_pred <- predict(lm_cub_model)

r2_lm_cub <- summary(lm_cub_model)$r.squared
##summary(lm_cub_model)
formula_lm_cub = paste(s_life_exp,"~", round(coefficients(lm_cub_model)[1], 2), 
                      "+\n    ", round(coefficients(lm_cub_model)[2], 2), "*", s_lngdp, 
                      "+\n    ", round(coefficients(lm_cub_model)[3], 2), "*", s_lngdp, 
                      "+\n    ", round(coefficients(lm_cub_model)[4], 2), "*", s_lngdp, sep = " ")

```


Visualizing the results:  



```{r model_viz, echo = FALSE, fig.height = 15}
l <- labs(x = s_gdp, y = s_life_exp)
s_y <- scale_y_continuous(limits = c(45, 90))
s_x <-scale_x_continuous(expand = c(0, 0), limits = c(0, 13500))

p_lowess <- ggplot(data = dt, aes(gdp, life_exp)) + gp + l + s_y + s_x +
  geom_line(data = dt, aes(gdp, lowess_pred), color = 'blue') +
  ggtitle("Model: Lowess") +
  geom_text(aes(x = 200, y = 86), label = paste("R^2 =", round(r2_lowess, 4), sep=" "), size = 4, hjust = 0)

p_lm <- ggplot(data = dt, aes(gdp, life_exp)) + gp + l + s_y + s_x +
  geom_line(data = dt, aes(gdp, lm_pred), color = 'blue') +
  ggtitle("Model: Linear Regression") +
  geom_text(aes(x = 200, y = 89), label = formula_lm, size = 4, hjust = 0) +
  geom_text(aes(x = 200, y = 86), label = paste("R^2 =", round(r2_lm, 4), sep =" ") , size = 4, hjust = 0)

p_lm_ln <- ggplot(data = dt, aes(gdp, life_exp)) + gp +  l + s_y + s_x +
  geom_line(data = dt, aes(gdp, lm_ln_pred), color = 'blue') +
  ggtitle("Model: Level - Log Linear Regression") +
  geom_text(aes(x = 200, y = 89), label = formula_lm_ln, size = 4, hjust = 0) +
  geom_text(aes(x = 200, y = 86), label = paste("R^2 =", round(r2_lm_ln, 4), sep =" "), size = 4, hjust = 0)

p_spline <- ggplot(data = dt, aes(gdp, life_exp)) + gp +  l + s_y + s_x +
  geom_line(data = dt, aes(gdp, spline_pred), color = 'blue', size = .7) +
  geom_vline(xintercept = exp(knot), color = 'red', size = .7) +
  ggtitle("Model: Spline w/ 1 knot") +
  geom_text(aes(x = 200, y = 87), label = formula_spline, size = 4, hjust = 0) +
  geom_text(aes(x = 200, y = 82), label = paste("R^2 =", round(r2_spline, 4), sep =" "), size = 4, hjust = 0)

p_lm_sq <- ggplot(data = dt, aes(gdp, life_exp)) + gp + l + s_y + s_x +
  geom_line(data = dt, aes(gdp, lm_sq_pred), color = 'blue', size = .7) +
  ggtitle("Model: Linear Regression w/ square polinom") +
  geom_text(aes(x = 200, y = 87), label = formula_lm_sq, size = 4, hjust = 0) +
  geom_text(aes(x = 200, y = 82), label = paste("R^2 =", round(r2_lm_sq, 4), sep =" "), size = 4, hjust = 0)

p_lm_cub <- ggplot(data = dt, aes(gdp, life_exp)) + gp + l + s_y + s_x +
  geom_line(data = dt, aes(gdp, lm_cub_pred), color = 'blue', size = .7) +
  ggtitle("Model: Linear Regression w/ cubic polinom") +
  geom_text(aes(x = 200, y = 85), label = formula_lm_cub, size = 4, hjust = 0) +
  geom_text(aes(x = 200, y = 79), label = paste("R^2 =", round(r2_lm_cub, 4), sep =" "), size = 4, hjust = 0)

multiplot(p_lowess, p_lm, p_spline, p_lm_ln, p_lm_sq, p_lm_cub, cols = 2)
```


Note on the visuals: The different plots has been scaled back to show the non-log GDP / capita on the X axis, even if the models were run on the log transformation.


The results follow our expactations - the more complex models tend to drive slightly better fits in the sample. Having small differences is actually key in selecting our favored model.

While a simple linear regression on non-log GDP is visibly not the best model, from models run on log(GDP per capita) we actually argue to select the simplest one, our log-linear regression. This model's fit in the sample (measured by R^2) is just insignificantly behind of the more complex models'. Also, based on p values, only the beta of log(GDP) variable is significant on the 95% and 99% confidence levels.

Supporting our decision, below is a summary for the different variables used in different models (level - level excluded):

```{r conf_int, echo = FALSE}
dt_summaries <- data.table(all.vars(as.formula(lm_ln_model)))
dt_summaries <- cbind(dt_summaries, data.table(summary(lm_ln_model)$coefficient))
dt_summaries <- cbind(dt_summaries, "Level - Log Linear Regression")

temp <- data.table(all.vars(as.formula(lm_model)))
temp <- cbind(temp, data.table(summary(lm_model)$coefficients))
temp <- cbind(temp, "Linear Regression")

dt_summaries <- rbind(dt_summaries, temp)

temp <- data.table(all.vars(as.formula(spline_model)))
temp <- cbind(temp, data.table(summary(spline_model)$coefficients))
temp <- cbind(temp, "Spline w/ 1 knot")

dt_summaries <- rbind(dt_summaries, temp)

temp <- data.table(all.vars(as.formula(lm_sq_model)))
temp <- cbind(temp, data.table(summary(lm_sq_model)$coefficients))
temp <- cbind(temp, "Linear Regression w/ square polinom")

dt_summaries <- rbind(dt_summaries, temp)

temp <- data.table(all.vars(as.formula(lm_cub_model)))
temp <- cbind(temp, data.table(summary(lm_cub_model)$coefficients))
temp <- cbind(temp, "Linear Regression w/ cubic polinom")

dt_summaries <- rbind(dt_summaries, temp)

setnames(dt_summaries, c('x', 'beta', 'se', 't', 'p', 'model'))

dt_summaries <- dt_summaries[x != "life_exp", ]

dt_summaries[, ci_low := beta - se]
dt_summaries[, ci_high := beta + se]

dt_summaries[, is_sign_95 := p < 0.05]
dt_summaries[, is_sign_99 := p < 0.01]

dt_summaries[, model := factor(model, 
                               levels = c("Level - Log Linear Regression", 
                                          "Linear Regression",
                                          "Spline w/ 1 knot", 
                                          "Linear Regression w/ square polinom", 
                                          "Linear Regression w/ cubic polinom")) ]

ggplot(dt_summaries[x == "lngdp", ], aes(x = x, y = beta, ymin = ci_low, ymax = ci_high)) + 
  geom_pointrange(color = 'orange') + 
  facet_grid(~model, scales="free")
```

```{r coeffs, echo = FALSE}
pander(dt_summaries, split.table = Inf)
```


The level-log linear model, with close to .7 R^2 actually carries a very strong predictive power. We can interpret the formula as follows: a country with 1% higher GDP / capita is expected to have a life expetancy higher by 0.0542 years on average.

One factor that this model leaves unexplored is the difference is different countries populations. We explore that in the next section.


#### 3. Estimate a weighted regression (weight=population). Compare results to what we saw in class.   


Modeling:  

```{r w_regression}
lm_weighted_model <- lm(life_exp ~ lngdp, weights = pop, data = dt)
dt$lm_weighted_pred <- predict(lm_weighted_model)
r2_lm_weighted <- summary(lm_weighted_model)$r.squared

formula_lm_weighted = paste(s_life_exp,"~", round(coefficients(lm_weighted_model)[1], 2), 
                      "+", round(coefficients(lm_weighted_model)[2], 2), "*", s_lngdp, "\nweighted by", s_pop, sep = " ")
```


Visualizing results:  


```{r model_viz_2, echo = FALSE, fig.height = 5, fig.width = 5}
ggplot(data = dt, aes(gdp, life_exp))  + gp + l + s_y + s_x +
  geom_line(data = dt, aes(gdp, lm_weighted_pred), color = 'blue', size = .7) +
  ggtitle("Model: Population weighted linear regression") +
  geom_text(aes(x = 200, y = 85), label = formula_lm_weighted, size = 4, hjust = 0) +
  geom_text(aes(x = 200, y = 81), label = paste("R^2 =", round(r2_lm_weighted, 4), sep =" "), size = 4, hjust = 0)
```


## TODO explain
