---
title: "DA3_2"
author: "Tamas Koncz"
date: '2017 november 18 '
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(data.table)
library(wbstats)
library(tidyverse)
library(lspline)


# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

#### Download cross-country data on life expectancy and GDP per capita. “GDP per capita, PPP (constant)” and “Life expectancy at birth (total)”

#### 1. Delete unnecessary columns and save a csv file with three columns only: country name, life expectancy and GDP per capita. Keep countries with non-missing values for life expectancy and GDP per capita. Document what you do. 

First, we'll use the WB API for getting our data set to R.


```{r data_collection}

x <- wb(country = "countries_only", indicator = c("NY.GDP.PCAP.PP.KD", "SP.DYN.LE00.IN", "SP.POP.TOTL") , startdate = 2015, enddate = 2015)
countries <- data.table(x)

```

Once we have the data loaded,format the data into a workable format.
We are making four transformations:

  1. We are creating separate columns for each of our variables
  
  2. Columns are renamed to short names for easier coding
  
  3. Rows with NA values are removed
  
  4. Population is redefined as population in millions for easier handling
  
As a last step, we save our data table to a csv file as requested in the exercise.

```{r data_prep_2}
dt <- countries[, c('value', 'country', 'indicator')]
dt <- spread(dt, key = 'indicator', value = 'value', fill = NA)  

#saving string names for later use (charts, ...)
s_country = colnames(dt)[1]
s_gdp = "GDP per capita, $" ##colnames(dt)[2]
s_life_exp = "Life exp. at birth, years)" ##colnames(dt)[3]
s_pop = "Population, million" ##colnames(dt)[4]
s_lnpop = "log(Population, million)"
s_lngdp = "log(GDP per capita, $)"

setnames(dt, old = colnames(dt), new = c('country', 'gdp', 'life_exp', 'pop'))

dt <- dt[is.na(gdp) == FALSE, ]
dt <- dt[is.na(pop) == FALSE, ]
dt <- dt[is.na(life_exp) == FALSE, ]

dt[, pop := pop / 1000000]
dt[, lngdp := log(gdp)]

write.csv(x = dt, file = 'WD_Data_Filtered.csv')
```

Before we create our models, let's explore our data set visually.
First, using histograms to visualize single-variable distributions:
```{r data_viz_1, error = FALSE, warning = FALSE, echo = FALSE, message = FALSE, fig.width = 10, fig.height = 3}
p1 <- ggplot(data = dt, aes(gdp)) + geom_histogram(binwidth = 2500) + 
  scale_x_continuous(labels = scales::comma) + 
  labs(x = s_gdp)
p2 <- ggplot(data = dt, aes(lngdp)) + geom_histogram(binwidth = .25) +
  labs(x = s_lngdp)

multiplot(p1, p2, cols = 2)

p3 <- ggplot(data = dt, aes(life_exp)) + geom_histogram(binwidth = 5) + 
  labs(x = s_life_exp)

p4 <- ggplot(data = dt, aes(pop)) + geom_histogram(binwidth = 25) + 
  scale_x_continuous(labels = scales::comma) + 
  labs(x = s_pop)

p5 <- ggplot(data = dt, aes(log(pop))) + geom_histogram() + 
  scale_x_continuous(labels = scales::comma) + 
  labs(x = paste("log(", s_pop, ")", sep = ""))

multiplot(p3, p4, p5, cols = 3)
```
  
  
Then, we'll take a look at multi-varite distributions with scatterplots (inc. default loess lines).
(Bubble sizes show the size of the population)  


```{r data_viz_2, error = FALSE, warning = FALSE, echo = FALSE, message = FALSE, fig.width = 10, fig.height = 6}
p6 <- ggplot(data = dt, aes(lngdp, life_exp)) + geom_point(size = log(dt$pop) / 1.5, color = 'orange') + 
  geom_smooth(method = 'loess') + 
  labs(x = s_lngdp, y = s_life_exp)

p8 <- ggplot(data = dt, aes(lngdp, pop)) + geom_point(size = log(dt$pop) / 1.5, color = 'orange') + 
  geom_smooth(method = 'loess') + 
  labs(x = s_lngdp, y = s_pop)

p7 <- ggplot(data = dt, aes(life_exp, pop)) + geom_point(size = log(dt$pop) / 1.5, color = 'orange') + 
  geom_smooth(method = 'loess') + 
  labs(x = s_life_exp, y = s_pop)

p9 <- ggplot(data = dt, aes(lngdp, log(pop))) + geom_point(size = log(dt$pop) / 1.5, color = 'orange') + 
  geom_smooth(method = 'loess') + 
  labs(x = s_lngdp, y = s_lnpop)

multiplot(p6, p7, p8, p9, cols = 2)
```



#### 2. Estimate a lowess regression of life expectancy on ln gdp per capita. Estimate a linear regression of life expectancy on GDP per capita that best captures the nonlinearity you found (life expectancy on a piecewise linear spline or a polynomial in the explanatory variable). Argue for your choice. Report the coefficient estimates as well as their confidence interval, interpret and visualize the results. 



